---
title: "Wisconsin Football Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T,eval=T,message=F,warning=F,fig.align='center')
library(tidyverse)
library(sqldf)
library(lemon)
library(kableExtra)
library(ggmap)
library(leaflet)
library(leaflet.providers)
library(glue)
library(geosphere)
```

**Helper functions**

```{r}
show_table = function(df, caption="") {
  return (df %>% kbl(caption = caption) %>% kable_classic_2(bootstrap_options = c("striped", "hover"), full_width = F, font_size=20, html_font="Cambria"))
}
```

## Setup

Read in player data from scraped `.csv` file.

```{r}
players_raw = read_csv("players.csv")


player_years = sqldf("SELECT Name, MIN(YEAR) AS FirstYear, MAX(YEAR) AS LastYear FROM players_raw GROUP BY Name")

players = sqldf("SELECT DISTINCT players_raw.Name, players_raw.Pos, players_raw.Hometown, player_years.FirstYear as FirstYear, player_years.LastYear as LastYear FROM players_raw LEFT JOIN player_years ON players_raw.Name=player_years.Name")
```

This removed all duplicates, while also keeping for what years those players played. However there are still some duplicates because a player could have switched positions between seasons. See this in the case of FB/OLB/TE Riley Nowakowski. We still have three individual rows for this one player.

```{r}
players %>% 
  filter(Name == "Riley Nowakowski") %>% 
  show_table()
```


```{r}
agg = aggregate(Pos~Name, players, FUN=c)

towns = c()
firstyears = c()
lastyears = c()

for (i in 1:length(agg$Name)) {
  player = head(players[players$Name == agg$Name[i],], 1)
  towns = append(towns, player$Hometown[1])
  firstyears = append(firstyears, player$FirstYear[1])
  lastyears = append(lastyears, player$LastYear[1])
  
}
    
players_no_dup = agg %>% 
  mutate(Hometown = towns, FirstYear = firstyears, LastYear = lastyears)

# Individual town adjustment
players_no_dup$Hometown[players_no_dup$Hometown == "Montreal, Wis."] = "Montreal, Wisconsin"
players_no_dup$Hometown[players_no_dup$Hometown == "Howard, Kan."] = "Howard, Kansas"

players_w_mult_pos = (nrow(players)-nrow(players_no_dup))/nrow(players) * 100

tibble("Pos. Not Combined" = nrow(players), "Pos. Combined" = nrow(players_no_dup), "% of Multiple Position" = round(players_w_mult_pos, 2)) %>% 
  show_table("Length of Players DF")
```

> After this cell, we see that the length of the players DataFrame has decreased significantly, showing that we correctly identified players with multiple positions, and combined them to one row.We also showed that we have `r round(players_w_mult_pos, 2)`% of players that have played (officially) multiple positions throughout their careers at Wisconsin.


## Hometowns

The first part of this analysis will be geographical. Where are Wisconsin football players originating from? **How** has that trend changed over the past decade?

### Graph Hometowns

I created a [script](./geocodes.R) to map every Hometown to its longitude and latitude geocode. The main reason for doing this was so there wouldn't be a ton of calls to the Google Maps API every time I ran this `.rmd` file.

```{r, include=F}
readRenviron("./.Renviron")
register_google(Sys.getenv("MAPSAPIKEY"))
```


```{r}
geocodes = read_csv("geocodes.csv")
```

```{r}
print_players = function(df) {
  string = ""
  if (is.na(df) || nrow(df) == 0){
   return (NA)
  }
  for (i in 1:nrow(df)){
    row = df[i,]
    player = ifelse(row$FirstYear == row$LastYear, glue('{row$Name}, {row$FirstYear}'), glue('{row$Name}, {row$FirstYear}-{row$LastYear}'))
    string = paste(string, player, "<br>")
  }
  return (string)
}
```

```{r}
town_players = c()
for (town in geocodes$Hometown) {
  town_players = append(town_players, list(players_no_dup[players_no_dup["Hometown"] == town,]))
  
}

geocodes$Players = town_players
```

```{r}
m = leaflet() %>% 
  addProviderTiles(provider=providers$Esri.WorldGrayCanvas) %>% 
  addCircleMarkers(geocodes$lon, geocodes$lat, popup=paste("<b>", geocodes$Hometown, "</b><br>", lapply(geocodes$Players, print_players)), label=geocodes$Hometown, radius=4.5, color="#C5050C", fillOpacity=1.0, stroke=FALSE)
m
```

Getting distance from Madison, WI for all cities in our dataset.
```{r}

year_dists = players_raw
# Fix the two individual cities
players_raw$Hometown[year_dists$Hometown == "Howard, Kan."] = "Howard, Kansas"
players_raw$Hometown[year_dists$Hometown == "Montreal, Wis."] = "Montreal, Wisconsin"
madison_dists = c()
madison = geocode("Madison WI")
for (i in 1:nrow(players_raw)){
  city = geocodes %>% filter(Hometown == players_raw$Hometown[i]) %>% select(lon, lat)
  if (is.na(city$lon[1]) || is.na(city$lat[1])){
    print(players_raw$Hometown[i])
  }
  madison_dists = append(madison_dists, distHaversine(madison, c(city$lon[1], city$lat[1])))
}
year_dists$dist_madison = madison_dists
```

Average Distance From Madison by Year
```{r}
avg_distances = year_dists %>% 
  group_by(Year) %>% 
  summarize(avg_dist = mean(dist_madison/1000))

ggplot(avg_distances, aes(Year, avg_dist)) +
  geom_col(fill="lightblue", color="black") +
  ylab("Distance (km)") +
  ggtitle("Average Distance from Madison of UW Football Players")
```



